Symbolic Weighted Language Models and\\ Parsing over Infinite Alphabets

### 2021-06-08 main.tex
rev. def. swTA (check parent node) and conversiion VPA

\def\SA{\textsf{sA}\xspace}
\def\WA{\textsf{wA}\xspace}
\def\SWT{\textsf{swT}\xspace}
\def\SWA{\textsf{swA}\xspace}
\def\SWTA{\textsf{swTA}\xspace}
\def\SWVPA{\textsf{sw-VPA}\xspace}
\def\weight{\mathsf{weight}}
\def\wei{\mathsf{w}}
\def\mei{\mathsf{m}}
\def\init{\mathsf{in}}
\def\final{\mathsf{out}}
\newcommand{\call}[1]{\ensuremath #1} %{\ensuremath \langle_{#1}}
\newcommand{\return}[1]{\ensuremath #1} %{\ensuremath {}_{#1}{\rangle}} % $\prescript{}{a}{)}$
\def\Omegai{{\Omega_\mathsf{i}}}
\def\Omegac{{\Omega_\mathsf{c}}}
\def\Omegar{{\Omega_\mathsf{r}}}
\def\Sigmai{{\Sigma_\mathsf{i}}}
\def\Sigmac{{\Sigma_\mathsf{c}}}
\def\Sigmar{{\Sigma_\mathsf{r}}}
\def\Deltai{{\Delta_\mathsf{i}}}
\def\Deltac{{\Delta_\mathsf{c}}}
\def\Deltar{{\Delta_\mathsf{r}}}
\def\Phii{{\Phi_\mathsf{i}}}
\def\Phic{{\Phi_\mathsf{c}}}
\def\Phir{{\Phi_\mathsf{r}}}
\def\Phix{{\Phi_\mathsf{x}}}
\def\Phici{{\Phi_\mathsf{ci}}}
\def\Phicc{{\Phi_\mathsf{cc}}}
\def\Phicr{{\Phi_\mathsf{cr}}}
\def\Phicx{{\Phi_\mathsf{cx}}}
\def\weii{{\wei_\mathsf{i}}}
\def\weic{{\wei_\mathsf{c}}}
\def\weir{{\wei_\mathsf{r}}}
\def\weie{{\wei_\mathsf{e}}}
\def\weiei{{\wei^\mathsf{e}_\mathsf{i}}}
\def\weiec{{\wei^\mathsf{e}_\mathsf{c}}}
\def\weier{{\wei^\mathsf{e}_\mathsf{r}}}
\def\weiex{{\wei^\mathsf{e}_\mathsf{x}}}
\newcommand{\ioi}[1]{\mathsf{ioi}({#1})}
\newcommand{\rank}{\mathsf{rk}}
\newcommand{\lin}{\mathsf{lin}}

@@

% A \emph{regular tree grammar} over $\Omega$ 
% is a triplet $G = \< N, q_\mathsf{i}, R>$ where
% $N$ is a finite set of non-terminal symbols denoted $q$..., 
% $q_\mathsf{i} \in N$ is the starting non-terminal, 
% $R$ is a finite set of production rules of the form
% $q_0 \to a(q_1\ldots q_k)$ where 
% $q_0, q_1, \ldots, q_k \in N$
% $a \in \Omega_k$.
% A tree $t \in \T(\Omega)$ is in the language of $G$ 
% if it can be generated from $q_\mathsf{i}$ by 
% non terminal replacement following the rules of $R$.
%
\begin{definition}  \label{def:SWTA}
A \emph{symbolic-weighted tree automaton} (\SWTA)
over $\Omega$, $\Semiring$, and $\bar\Phi$
is a triplet $A = \< Q, \init, \bar{\wei} >$ where
$Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$ is the starting weight function, 
and $\bar{\wei}$ is a tuplet of transition functions containing
%$\wei_\varepsilon$ from $Q \times Q$ into $\Semiring$, and, 
one function $\wei_{k}: Q \times Q^{k} \to \Phi_{\Omega_k}$
for each $k \in [0..M]$. 
\end{definition}
%
%Like in Section~\ref{sec:SWAdef}, 
We define from $\bar{\wei}$ 
a transition function~$\wei$, %from $\bar\wei$
from 
$Q \times \Omega \times \bigcup_{k=0}^{M} Q^k$
into~$\Semiring$ by: %also called $\wei$ for simplicity, 
%such that, for all $q, q' \in Q$, $a \in \Sigma$, and $b \in \Delta$, 
\[
\begin{array}{rcll}
%\wei(q_0, \varepsilon, q_1) & = &  \wei_\varepsilon(q_0, q_1),\\ %\phi_\varepsilon\\
\wei(q_0, a, q_1 \ldots q_k) & = & \phi_{k}(a) &
\quad\mathrm{where~} \phi_{k} = \wei_{k}(q_0, q_1\ldots q_k).\\
\end{array}      
\]
%
\noindent
The \SWTA of Definition~\ref{def:SWTA} defines a mapping 
from trees of $\T(\Omega)$ into the weight values in~$\Semiring$,
based on the intermediate function~$\weight_A$
defined as follows for $q_0 \in Q$ and 
$t = b(t_1,\ldots, t_k) \in \T(\Omega)$,
with $0 \leq k \leq M$:
\begin{equation}
\begin{array}{rccl}
\weight_A(q_0, t) & = & 
%   \displaystyle\bigoplus_{q_1 \in Q} &
%   \wei(q, \varepsilon, q_1) \otimes \weight_A(q_1, t)\\
 \displaystyle\bigoplus_{q_1 \ldots q_k \in Q^k} &
              \wei(q_0, b, q_1 \ldots q_k ) 
   \otimes \displaystyle\bigotimes_{i=1}^{k}
           \weight_A(q_i, t_i)
\end{array}
\end{equation}

\medskip\noindent
The weight associated by $A$ to  $t \in \T(\Omega)$ is 
\begin{equation}
A(t)  = 
\displaystyle\bigoplus_{q \in Q} \mathsf{in}(q) \mathop{\otimes} \weight_A(q, t)
\label{eq:weightTA}
\end{equation}

\noindent
Intuitively, $\wei(q_0, a, q_1 \ldots q_k)$ can be seen as
the weight of a production rule $q_0 \to a(q_1, \ldots, q_k)$ 
of a regular tree grammar $G$~\cite{tata}, 
that replaces the non-terminal symbol $q_0$ by $a(q_1, \ldots, q_k)$. 
%in a step of tree building.
%
%Such a grammar computes the weights of the derivation trees 
%of the Context-Free grammar obtained by forgetting the labeling symbols of $\Omega_{>0}$.
And the above production rule can also be seen as 
a rule of a weighted CF grammar, of the form
$[a]\, q_0 := q_1 \ldots q_k$ if $k > 0$,
and $q_0 := a$ if $k = 0$. 
In the first case, $a$ is a label of the rule, 
in the second case it is a terminal symbol.
On can observe that the weight of a labeled derivation tree $t$
of the weighted CF grammar associated to~$A$ as above, 
is $\weight_A(q, t)$, 
when $q$ is the start non-terminal.
%
We shall now establish a correspondence between such derivation tree $t$
and some word describing a linearization of $t$, 
in a way that $\weight_A(q, t)$ can be computed by a $\SWVPA$.

@@

\[
\begin{array}{ll}
\weii(\bar{q}, {\call{b}}, \bar{p}, a, \bar{q} q') = \wei(q', a, \varepsilon) & 
\mathrm{for~all~} b \in \Omega_{>0}, \bar{p}\in P', a \in \Omega_0\\
%
\weiei(\bar{q}, a, \bar{q} q') = \wei(q', a, \varepsilon) & 
\mathrm{for~all~} a \in \Omega_0\\
%
\weic(\bar{q}, {\call{b}}, \bar{p}, \call{b'}, \varepsilon, \bar{q}) = \one & %\wei(p', b, \bar{p}) & 
\mathrm{for~all~} b \in \Omega_{>0},  \bar{p}\in P', b, b' \in \Omega_{>0}\\
%
\weiec(\bar{q}, \call{b}, \varepsilon, \bar{q}) = \one & %\wei(p', b, \bar{p}) & 
\mathrm{for~all~} b \in \Omega_{>0}\\
%
%\weir: Q \times \Omegac \times P \times \Sigmar \times Q \to \Semiring & 
\weir(\bar{q}, {\call{b}}, \bar{p}, {\return{b}}, \bar{p} q') = \wei(q', b, \bar{q}) & 
\mathrm{for~all~}  b \in \Omega_{>0}, \bar{p}\in P'\\
%
%\weie: Q \times \Omegar \times Q \to \Semiring & 
\weier(\bar{p}, {\return{b}}, \bar{q}) = \zero &
\mathrm{for~all~}  b \in \Omega_{>0}\\
\end{array}      
\]
\noindent
It is sufficient to consider in $Q'$ only the prefixes of 
sequences in transition with a non-null weight.
\qed
\end{proof}

>>>>

\def\SA{\textsf{sA}\xspace}
\def\WA{\textsf{wA}\xspace}
\def\SWT{\textsf{swT}\xspace}
\def\SWA{\textsf{swA}\xspace}
\def\SWTA{\textsf{swTA}\xspace}
\def\SWVPA{\textsf{sw-VPA}\xspace}
\def\weight{\mathsf{weight}}
\def\wei{\mathsf{w}}
\def\mei{\mathsf{m}}
\def\init{\mathsf{in}}
\def\final{\mathsf{out}}
\newcommand{\call}[1]{\ensuremath #1} %{\ensuremath \langle_{#1}}
\newcommand{\return}[1]{\ensuremath #1} %{\ensuremath {}_{#1}{\rangle}} % $\prescript{}{a}{)}$
\def\Omegai{{\Omega_\mathsf{i}}}
\def\Omegac{{\Omega_\mathsf{c}}}
\def\Omegar{{\Omega_\mathsf{r}}}
\def\Sigmai{{\Sigma_\mathsf{i}}}
\def\Sigmac{{\Sigma_\mathsf{c}}}
\def\Sigmar{{\Sigma_\mathsf{r}}}
\def\Deltai{{\Delta_\mathsf{i}}}
\def\Deltac{{\Delta_\mathsf{c}}}
\def\Deltar{{\Delta_\mathsf{r}}}
\def\Phii{{\Phi_\mathsf{i}}}
\def\Phic{{\Phi_\mathsf{c}}}
\def\Phir{{\Phi_\mathsf{r}}}
\def\Phix{{\Phi_\mathsf{x}}}
\def\Phici{{\Phi_\mathsf{ci}}}
\def\Phicc{{\Phi_\mathsf{cc}}}
\def\Phicr{{\Phi_\mathsf{cr}}}
\def\Phicx{{\Phi_\mathsf{cx}}}
\def\weii{{\wei_\mathsf{i}}}
\def\weic{{\wei_\mathsf{c}}}
\def\weir{{\wei_\mathsf{r}}}
\newcommand{\weie[1]}{\wei_{#1}^\mathsf{e}}
\def\weiei{\weie[\mathsf{i}]}
\def\weiec{\weie[\mathsf{c}]}
\def\weier{\weie[\mathsf{r}]}
\def\weiex{\weie[\mathsf{x}]}
\newcommand{\ioi}[1]{\mathsf{ioi}({#1})}
\newcommand{\rank}{\mathsf{rk}}
\newcommand{\lin}{\mathsf{lin}}

@@

% A \emph{regular tree grammar} over $\Omega$ 
% is a triplet $G = \< N, q_\mathsf{i}, R>$ where
% $N$ is a finite set of non-terminal symbols denoted $q$..., 
% $q_\mathsf{i} \in N$ is the starting non-terminal, 
% $R$ is a finite set of production rules of the form
% $q_0 \to a(q_1\ldots q_k)$ where 
% $q_0, q_1, \ldots, q_k \in N$
% $a \in \Omega_k$.
% A tree $t \in \T(\Omega)$ is in the language of $G$ 
% if it can be generated from $q_\mathsf{i}$ by 
% non terminal replacement following the rules of $R$.
%
\begin{definition}  \label{def:SWTA}
A \emph{symbolic-weighted tree automaton} (\SWTA)
over $\Omega$, $\Semiring$, and~$\bar\Phi$
is a triplet $A = \< Q, \init, \bar{\wei} >$ where
$Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Phi_\Omega$ is the starting weight function, 
and $\bar{\wei}$ is a tuplet of transition functions containing, 
for each $k \in [0..M]$, 
%$\wei_\varepsilon$ from $Q \times Q$ into $\Semiring$, and, 
the functions $\wei_{k}: Q \times Q^{k} \to \Phi_{\Omega_{>0},\Omega_k}$
and $\weie[k]: Q \times Q^{k} \to \Phi_{\Omega_k}$.
\end{definition}
%
%Like in Section~\ref{sec:SWAdef}, 
We define %from $\bar{\wei}$ 
a transition 
function~$\wei: Q \times (\Omega_{> 0} \cup \{ \varepsilon \}) \times \Omega \times \bigcup_{k=0}^{M} Q^k 
  \to \Semiring$
by: %also called $\wei$ for simplicity, 
%such that, for all $q, q' \in Q$, $a \in \Sigma$, and $b \in \Delta$, 
\[
\begin{array}{rcll}
%\wei(q_0, \varepsilon, q_1) & = &  \wei_\varepsilon(q_0, q_1),\\ %\phi_\varepsilon\\
\wei(q_0, a, b, q_1 \ldots q_k) & = & \eta(a, b) &
\quad\mathrm{where~} \eta = \wei_{k}(q_0, q_1\ldots q_k)\\
\wei(q_0, \varepsilon, b, q_1 \ldots q_k) & = & \phi(b) &
\quad\mathrm{where~} \phi = \weie[k](q_0, q_1\ldots q_k).
\end{array}      
\]
%
\noindent
The first case deals with a strict subtree, with a parent node labeled by $a$, 
and the second case with a root tree.
%
Every \SWTA %of Definition~\ref{def:SWTA} 
defines a mapping 
from trees of $\T(\Omega)$ into the weight values in~$\Semiring$,
based on an intermediate function
$\weight_A: Q \times (\Omega \cup \{ \varepsilon \}) \times \T(\Omega) \to \Semiring$
defined as follows 
\begin{equation}
\begin{array}{rccl}
\weight_A(q_0, a, t) & = & 
%   \displaystyle\bigoplus_{q_1 \in Q} &
%   \wei(q, \varepsilon, q_1) \otimes \weight_A(q_1, t)\\
 \displaystyle\bigoplus_{q_1 \ldots q_k \in Q^k} &
              \wei(q_0, a, b, q_1 \ldots q_k ) 
   \otimes \displaystyle\bigotimes_{i=1}^{k}
           \weight_A(q_i, b, t_i)
\end{array}
\end{equation}
where $q_0 \in Q$, $a \in \Omega_{>0} \cup \{ \varepsilon \}$ and 
$t = b(t_1,\ldots, t_k) \in \T(\Omega)$,
$0 \leq k \leq M$.

\medskip\noindent
Finally, the weight associated by $A$ to  $t \in \T(\Omega)$ is 
\begin{equation}
A(t)  = 
\displaystyle\bigoplus_{q \in Q} \mathsf{in}(q) \mathop{\otimes} \weight_A(q, \varepsilon, t)
\label{eq:weightTA}
\end{equation}

\noindent
Intuitively, $\wei(q_0, a, b, q_1 \ldots q_k)$ can be seen as
the weight of a production rule $q_0 \to b(q_1, \ldots, q_k)$ 
of a regular tree grammar~\cite{tata}, 
that replaces the non-terminal symbol $q_0$ by $b(q_1, \ldots, q_k)$, 
provided that the parent of $q_0$ is labeled by $a$
(or $q_0$ is the root node if $a = \varepsilon$).
%in a step of tree building.
%
%Such a grammar computes the weights of the derivation trees 
%of the Context-Free grammar obtained by forgetting the labeling symbols of $\Omega_{>0}$.
The above production rule can also be seen as 
a rule of a weighted CF grammar, of the form
$[a, b]\, q_0 := q_1 \ldots q_k$ if $k > 0$,
and $[a]\, q_0 := b$ if $k = 0$. 
In the first case, $b$ is a label of the rule, 
and in the second case, it is a terminal symbol.
And in both cases, $a$ is a constraint on the label of the 
parent node in the derivation tree.
This features of observation of parent's label  
are useful in the case of infinite alphabet, 
where it is not possible to memorize a label with the states.
%
%One can observe that 
The weight of a labeled derivation tree $t$
of the weighted CF grammar associated to~$A$ as above, 
is $\weight_A(q, t)$, 
when $q$ is the start non-terminal.
%
We shall now establish a correspondence between such derivation tree $t$
and some word describing a linearization of $t$, 
in a way that $\weight_A(q, t)$ can be computed by a $\SWVPA$.

@@

\[
\begin{array}{ll}
\weii(\bar{q}, {\call{c}}, \bar{p}, a, \bar{q} q') = \wei(q', c, a, \varepsilon) & 
\mathrm{for~all~} c \in \Omega_{>0}, \bar{p}\in P', a \in \Omega_0\\
%
\weiei(\bar{q}, a, \bar{q} q') = \wei(q', \varepsilon, a, \varepsilon) & 
\mathrm{for~all~} a \in \Omega_0\\
%
\weic(\bar{q}, {\call{c}}, \bar{p}, \call{c'}, \varepsilon, \bar{q}) = \wei(q', c, c', \bar{q}) &  
\mathrm{for~all~} c, c' \in \Omega_{>0},  \bar{p}\in P'\\
%
\weiec(\bar{q}, \call{c}, \varepsilon, \bar{q}) = \wei(q', \varepsilon, c, \bar{q}) & 
\mathrm{for~all~} c \in \Omega_{>0}\\
%
%\weir: Q \times \Omegac \times P \times \Sigmar \times Q \to \Semiring & 
\weir(\bar{q}, {\call{b}}, \bar{p}, {\return{b}}, \bar{p} q') = \one & 
\mathrm{for~all~}  b \in \Omega_{>0}, \bar{p}\in P'\\
%
%\weie: Q \times \Omegar \times Q \to \Semiring & 
\weier(\bar{p}, {\return{b}}, \bar{q}) = \zero &
\mathrm{for~all~}  b \in \Omega_{>0}\\
\end{array}      
\]
\noindent
It is sufficient to consider in $Q'$ only the prefixes of 
sequences in transition with a non-null weight.
\qed
\end{proof}
